##Part 1 - Loading and preprocessing the data
```{r,echo=FALSE}
setwd("C:\\Users\\Fawcett\\Documents\\Coursera\\R Working Directory\\Reproducible\\Assignment_1")
```

ggplot2, plyr and dplyr packages loaded - to be used later
```{r}
library(ggplot2)
library(plyr)
library(dplyr)
```

The data was read into R using the read.csv function.  The date column was then converted to the date format.  Note:  the csv file needs to be in the working directory.
```{r}
#read the data
step_data<-read.csv("activity.csv",header=TRUE,na.strings="NA")

#Format the dates
step_data$date<-as.Date(step_data$date)
str(step_data)
```



##Part 2 - What is mean total number of steps taken per day?

Calculate the total steps each day using functions from the dplyr package
```{r}
#Calculate the steps per day
steps_per_day<-step_data %>%
        na.omit() %>%
        group_by(date) %>%
        summarize(total_steps=sum(steps))
steps_per_day

```

Create a histogram using the base plotting system
```{r}
hist(steps_per_day$total_steps,breaks=20,xlab="Total Steps per Day",
     main="Histogram of Total Steps per Day")
```

Calculate the Mean and Median steps per day
```{r}
mean_steps<-format(mean(steps_per_day$total_steps),big.mark=",",scientific=FALSE)

median_steps<-format(median(steps_per_day$total_steps),big.mark=",",scientific=FALSE)

```

The Mean number of steps per day is `r mean_steps` and the Median is `r median_steps`.



##Part 3 - What is the average daily activity pattern?


Calculate the average number of steps taken per time period - averaged across all days
```{r}

step_intervals<-step_data %>%
        na.omit() %>%
        group_by(interval) %>%
        summarize(average_steps=mean(steps))
```

Plot the results
```{r}
plot(step_intervals,type="l",xlab="5-Minute Interval",
     ylab="Average Number of Steps Taken",
     main="Average Number of Steps Taken per 5-Minute Interval:  
     Averaged Accross All Days")
```

Calculate the 5-Minute Interval with the maximum average number of steps
```{r}
maximum_steps<-max(step_intervals$average_steps)
maximum_interval<-step_intervals$interval[step_intervals$average_steps==maximum_steps]
```
The interval with the highest number of steps is interval #`r maximum_interval` with an average of `r round(maximum_steps,2)` steps.



##Part 4 - Imputing missing values

Calculate the number of missing values in each column
```{r}
apply(is.na(step_data),2,sum)
```

Calculte the number of missing values and the proportion of missing values for the "steps" column.
```{r}
missing_value_count<-sum(is.na(step_data$steps))
missing_value_prop<-mean(is.na(step_data$steps))
```
There are `r missing_value_count` missing values, which is `r round(missing_value_prop * 100,2)`% of the total rows.

Fill in missing data with the average of the 5-Minute Interval
```{r}
imputed_data<-step_data

missing_data_index<-which(is.na(imputed_data$steps))

for (i in missing_data_index) {
        imputed_data$steps[i]<-round(with(step_intervals,average_steps[interval==imputed_data$interval[i]]),0)
}
```

Calculate the total steps each day using functions from the dplyr package
```{r}
imputed_steps_per_day<-imputed_data %>%
        group_by(date) %>%
        summarize(total_steps=sum(steps))
```

Create a histogram using the base plotting system
```{r}
par(mfrow=c(1,2))
hist(steps_per_day$total_steps,breaks=20,ylim=c(0,18),xlab="Total Steps per Day",
     main="Histogram of Total Steps per Day
     Raw")
hist(imputed_steps_per_day$total_steps,breaks=20,ylim=c(0,18),xlab="Total Steps per Day",
     main="Histogram of Total Steps per Day
     Imputed")
```

Calculate the Mean and Median steps per day
```{r}
imputed_mean_steps<-format(mean(imputed_steps_per_day$total_steps),big.mark=",",scientific=FALSE)

imputed_median_steps<-format(median(imputed_steps_per_day$total_steps),big.mark=",",scientific=FALSE)

```

With the imputed values the Mean number of steps per day is `r imputed_mean_steps` and the Median is `r imputed_median_steps`.  The mean is 0.55 steps per day lower than with the raw data, and the median went down 3 steps per day.


##Part 5 - Are there differences in activity patterns between weekdays and weekends?

Create a factor variable with "Weekday" and "Weekend"
```{r}
imputed_data2<-imputed_data

imputed_data2[weekdays(imputed_data2$date)%in%c("Saturday","Sunday"),"day_type"]<-"Weekend"
imputed_data2[is.na(imputed_data2$day_type),"day_type"]<-"Weekday"
imputed_data2$day_type<-factor(imputed_data2$day_type)
```

Plot the averge steps per interval - divided into Weekday and Weekend
```{r}
imputed_step_intervals2<-imputed_data2 %>%
        group_by(interval,day_type) %>%
        summarize(average_steps=mean(steps))

g<-ggplot(imputed_step_intervals2, aes(x = interval, y = average_steps))

g +
        geom_line(col = 'Blue') + 
        ylab('Average number of steps') + 
        xlab('interval') + 
        ggtitle('Average steps per interval') + 
        facet_grid(day_type ~ .)
```